---
title: 'MPP-E1180 Lecture 5: Intro to Markup Lang. & Literate Programming (2)'
author: "Christopher Gandrud"
date: "9 October 2014"
bibliography:  
            - RpackageCitations.bib
            - MainLit.bib
output:
  ioslides_presentation:
    css: http://maxcdn.bootstrapcdn.com/font-awesome/4.2.0/css/font-awesome.min.css
    logo: https://raw.githubusercontent.com/christophergandrud/Hertie_Collab_Data_Science/master/img/HertieCollaborativeDataLogo_v1.png
  beamer_presentation: default
---

## <i class="fa fa-arrow-circle-o-up"></i> Objectives for the week

- Housecleaning

- Review

- Advanced topics in markup languages and literate programming.

    + R Markdown Headers

    + BibTeX/Pandoc and citing R Packages

    + Time consuming analyses: caching and Make files

- Begin working on your Assignment 2.

## <i class="fa fa-check"></i> Class reschedule

Result of voting:

[GET]

## Assignment 2

**Proposal** for your Collaborative Research Project.

**Deadline:** Week 6

**Submbit:** A (max) 2,000 word proposal created with **RMarkdown**. The
proposal will:

- State your research question. And justify why it is interesting.

- Provide a basic literature review (properly cited with BibTeX).

- Identify data sources and appropriate research methodologies for answering your
question.

As always, submit the entire GitHub repo.

## Assignment 2

<br>
<br>
<br>

**Definitely see me with your ideas/draft.**

Start thinking about types of models that you want to use. I can include these
in Lecture 8 (Statistical Modeling with R).

## <i class="fa fa-arrow-left"></i> Review

- Why is literate programming useful for reproducible research?

- What is a markup language?

- What is a code chunk?

- What is the difference between Weave, Sweave, and knit? (kind of a trick question)

## R Markdown Headers

An R Markdown file is **just a text file** with markup instructions that **RStudio**
understands.

The key to formatting is the **header**.

It is at the start of a file and comes between `---`.

The header is written in [YAML](http://en.wikipedia.org/wiki/YAML).

## YAML

YAML is basically a human read-able data format.

Elements are separated from values with a colon (`:`).

Each element is separated by new lines.

Hierarchy is maintained with tabs (very different from R where tabs are meaningless).

<br>

Super Nerd Point: YAML is a recursive acronym: ''YAML Ain't Markup Language''.

## Example: Different Presentation Styles

[FILL]

## BibTeX citations

BibTex allows you to create a single data base of all of the literature/packages
you cite.

You can then insert them into your text and they will:

- Be automatically formatted consistently.

- Generate an appropriately ordered, consistently formatted reference list at
the end of your document.

## The BibTeX Database

A BibTeX database is just a text file with the extension `.bib`.

Each entry follows a standard format depending on the type of media.

```
@DOCUMENT_TYPE{CITE_KEY,
    title = {TITLE},
    author = {AUTHOR},
    . . . = {. . .},
}
```

**Note**: Commas are very important!

## The Cite Key

<br>
<br>
<br>

The cite key **links** a specific citation in your presentation document to a specific
BibTeX database entry.

They must be **unique**.

It **does not matter** what order your BibTeX entries are in the `.bib` file.

## BibTeX Articles

```
@article{Acemoglu2000,
    author = {Acemoglu, Daron and Robinson, James A.},
    title = {Why Did the West Extend the Franchise? Democracy, Inequality,
            and Growth in Historical Perspective},
    journal = {The Quarterly Journal of Economics},
    year = {2000},
    volume = {115},
    number = {4},
    pages = {1167--1199},
}
```

## BibTeX Books

```
@book{Cox1997,
    title={Making Votes Count: Strategic Coordination in the World's
            Electoral Systems},
    author={Gary W. Cox},
    year={1997},
    publisher={Cambridge University Press},
    address = {Cambridge}
}
```

## More

<br>
<br>

For more media types and entry fields see
[http://en.wikipedia.org/wiki/BibTeX](http://en.wikipedia.org/wiki/BibTeX).

## Tip: Google Scholar

Google scholar generates BibTeX entries.

On an entry click `Cite > BibTeX`.

Sometimes they need to be **cleaned** a little.

## Linking your .bib file.

To link your `.bib` file to your RMarkdown document add:

```
bibliography:
    - BIB_FILE_NAME.bib
    - ANOTHER_BIB_FILE_NAME.bib
```

**Note**: The files should be in the same directory as you R Markdown file.

## Including BibTeX citations in RMarkdown

RMarkdown uses [Pandoc syntax](http://johnmacfarlane.net/pandoc/README.html)
to include a citation in text.

General format: `@CITE_KEY`.

So if the cite key is `Box1973` then `@Box1973` will return  @Box1973 in the text
of the presentation document.

## Formatting In-Text Citations

| Markup                      | Result                   |
| --------------------------- | ------------------------ |
| `[@Box1973]`                | [@Box1973]               |
| `[see @Box1973]`            | [see @Box1973]           |
| `[see @Box1973, 33-40]`     | [see @Box1973, 33-40]    |
| `[@Box1973; @Acemoglu2000]` | [@Box1973 @Acemoglu2000] |
| `@Box1973 [33-40]`          | @Box1973 [33-40]         |

## Reference List

A reference list with the full bibliographic details of all cited documents
will be **automatically created** at the end of your document.

Tip: Put a `## References` at the very end of your R Markdown document to
have a section heading before the reference list.

## Citing R and R Packages

Why cite?

- Give **credit** to the package authors (just like when citing literature).

- Enable reproducible research: identify which software you used and **which
version**.

## Citing R and R Packages

Base R way: print citation, copy BibTeX  entry into your *.bib* file.

**Cite R**:

```{r}
toBibtex(citation())
```

## Citing R and R Packages

**Cite R Packages**:

```{r}
toBibtex(citation('dplyr'))
```

## Citing R and R Packages: LoadandCite

The dynamic literate programming way: Use `LoadandCite` from the
[repmis](http://christophergandrud.github.io/repmis/) package.

Load all of the packages at the beginning of you RMarkdown file in a chunk
with `include=FALSE`.

`LoadandCite` loads the packages and creates a BibTeX file with all of the citations.

```{r, message=FALSE, warning=FALSE}
pkgs <- c('dplyr', 'ggplot2')

repmis::LoadandCite(pkgs, file = 'RpackageCitations.bib')
```

**Note:** Use a file name that is different from your literature BibTeX file!

## Citing R and R Packages: LoadandCite

Include the .bib file in your RMarkdown header.

Each **cite key** follows: `R-PKG_NAME`.

R itself has the key `CiteR`.

So `@R-dplyr` and `@CiteR` create the citations:

- @R-dplyr

- @CiteR


## <i class="fa fa-clock-o"></i> Time Consuming/Intensive Analyses

Knitting your analysis and presentation documents together by placing all of your
R code into code chunks can sometimes be **problematic**:

- When they are time consuming (requires a lot of computational time).

- When they access files over the internet (bad practice to make many repeated
    calls to the same URL, can crash the site).

- When they are many lines long.

## Solutions

- Long lines: use `source()` to run R code in other files.

- `cache=TRUE` code chunk option: only runs the chunk when the **chunk code
changes**.

## Make files

Make files are the ultimate solution to these problems.

Make is a command line program.

Big Idea: run a *make* file that runs a list of specific files in order.

Files are only run if they have been changed since the last time the make
file was last run.

See me if you might want to do this.

## <i class="fa fa-arrow-circle-o-up"></i> Seminar: Begin working on your Proposal

Begin working with your partner on your research proposal.

- Identify the research area and key literature.

- Create a new repo and R Markdown document for your proposal.

- Begin building a BibTeX database for your key literature and try including them
in your proposal.

- Begin identifying data sources.

## References
