---
title: 'MPP-E1180 Lecture 5: Intro to Markup Lang. & Literate Programming (2)'
author: "Christopher Gandrud"
date: "25 September 2015"
output:
  beamer_presentation: default
  ioslides_presentation:
    css: https://maxcdn.bootstrapcdn.com/font-awesome/4.4.0/css/font-awesome.min.css
    logo: https://raw.githubusercontent.com/christophergandrud/Hertie_Collab_Data_Science/master/img/HertieCollaborativeDataLogo_v1.png
bibliography:
- RpackageCitations.bib
- MainLit.bib
---

## <i class="fa fa-arrow-circle-o-up"></i> Objectives for the topic

- Review

- Advanced topics in markup languages and literate programming.

    + R Markdown Headers

    + Footnotes

    + BibTeX/Pandoc and citing R Packages

    + Time consuming analyses: caching and Make files

## Assignment 2

**Proposal** for your Collaborative Research Project.

**Deadline:** Week 23

**Submbit:** A (max) 2,000 word proposal created with **R Markdown**. The
proposal will:

- State your research question. And justify why it is interesting.

- Provide a basic literature review (properly cited with BibTeX).

- Identify data sources and appropriate research methodologies for answering your
question.

As always, submit the entire GitHub repo.

## Assignment 2

<br>
<br>
<br>

**Definitely see me with your ideas/draft.**

Start thinking about **types of models** that you want to use. I can include these
in Lecture 8 (Statistical Modeling with R).

## <i class="fa fa-arrow-left"></i> Review

- Why is literate programming useful for reproducible research?

- What is a markup language?

- What is a code chunk?

- What is the difference between Weave, Sweave, knitr, and R Markdown? (kind of a trick question)

## Example Paper

An example of a paper + analysis + data project using many of the tools we cover
today is available at:

[HertieDataScience/Examples/SimplePaperWithAnalysis](https://github.com/HertieDataScience/Examples/tree/master/SimplePaperWithAnalysis)

## R Markdown Headers

An R Markdown file is **just a text file** with markup instructions that **RStudio**
understands.

The key to formatting is the **header**.

It is at the start of a file and comes between `---`.

The header is written in [YAML](http://en.wikipedia.org/wiki/YAML).

## YAML

YAML is a **human read-able data format**.

Elements are separated from values with a colon (`:`).

Each element is separated by new lines.

Hierarchy is maintained with tabs.

```
---
title: 'MPP-E1180 Lecture 5'
author: "Christopher Gandrud"
date: "9 October 2014"
output:
  ioslides_presentation:
    css: css/font-awesome.min.css
    logo: img/logo.png
  beamer_presentation: default
---
```

## Super Nerd Point

<br>
<br>
<br>

YAML is a recursive acronym: ''YAML Ain't Markup Language''.

## Different Presentation Styles

By defualt, R Markdown uses the [isoslides](http://rmarkdown.rstudio.com/ioslides_presentation_format.html)
HTML presentation slides style.

You can also use [reveal.js](http://lab.hakim.se/reveal-js/).

First install the revealjs R package:

```{r, eval=FALSE}
devtools::install_github("jjallaire/revealjs")
```

Then in the YAML header use:

```
output: revealjs::revealjs_presentation
```

For further styling see <https://github.com/jjallaire/revealjs>

## Table of contents & Numbered Sections

You can add a table of contents and numbered sections to your PDF output:

```
output:
  pdf_document:
    toc: true
    number_sections: true
```

To do the same for **HTML** also include the information under `html_document`.

## Figure Options

Create consistent figure formatting:

```
output:
  pdf_document:
    fig_width: 7
    fig_height: 6
    fig_caption: true
```

`fig_caption: true` attaches captions to figures.

To set the actual caption label, use the `fig.cap='SOME CAPTION'` **code chunk
option**.

## Pandoc footnotes

R Markdown can use Pandoc footnotes.

**In-text**: In the text place a **unique** footnote key in the format:

- `[^KEY]`

**At the end** of your document put the full footnote starting with the key,
e.g.

- `[^KEY]: This is a footnote.`

## BibTeX citations

BibTex allows you to create a **database** of **all** of the **literature/packages
you cite**.

You can then insert them into your text and they will:

- Be **automatically formatted** consistently.

- Generate an appropriately ordered, consistently formatted **reference list** at
the end of your document **with only the works you actually cited**.

## The BibTeX Database

A BibTeX database is just a text file with the extension `.bib`.

Each entry follows a standard format depending on the type of media.

```
@DOCUMENT_TYPE{CITE_KEY,
    title = {TITLE},
    author = {AUTHOR},
    . . . = {. . .},
}
```

**Note**: Commas are very important!

## The Cite Key

<br>
<br>
<br>

The cite key **links** a specific citation in your presentation document to a specific
BibTeX database entry.

They must be **unique**.

It **does not matter** what order your BibTeX entries are in the `.bib` file.

## BibTeX Articles

```
@article{Acemoglu2000,
    author = {Acemoglu, Daron and Robinson, James A.},
    title = {Why Did the West Extend the Franchise? Democracy, Inequality,
            and Growth in Historical Perspective},
    journal = {The Quarterly Journal of Economics},
    year = {2000},
    volume = {115},
    number = {4},
    pages = {1167--1199},
}
```

## BibTeX Books

```
@book{Cox1997,
    title={Making Votes Count: Strategic Coordination in the World's
            Electoral Systems},
    author={Gary W. Cox},
    year={1997},
    volume = {7},
    publisher={Cambridge University Press},
    address = {Cambridge}
}
```

## More

<br>
<br>

For more media types and entry fields see
[http://en.wikipedia.org/wiki/BibTeX](http://en.wikipedia.org/wiki/BibTeX).

## Tip: Google Scholar

Google scholar generates BibTeX entries.

On an entry click `Cite > BibTeX`.

For a YouTube how-to see [https://www.youtube.com/watch?v=SsJSR2b4_qc](https://www.youtube.com/watch?v=SsJSR2b4_qc).

Sometimes they need to be **cleaned** a little.

## Linking your .bib file.

To link your `.bib` file to your RMarkdown document add to the header:

```
bibliography:
    - BIB_FILE_NAME.bib
    - ANOTHER_BIB_FILE_NAME.bib
```

**Note**: The files should be in the **same directory** as your R Markdown file.

## Including BibTeX citations in RMarkdown

R Markdown uses [Pandoc syntax](http://johnmacfarlane.net/pandoc/README.html)
to include a citation in-text.

General format: `@CITE_KEY`.

So if the cite key is `Box1973` then `@Box1973` will return  @Box1973 in the text
of the presentation document.

## Formatting In-Text Citations

| Markup                      | Result                    |
| --------------------------- | ------------------------- |
| `[@Box1973]`                | [@Box1973]                |
| `[see @Box1973]`            | [see @Box1973]            |
| `[see @Box1973, 33-40]`     | [see @Box1973, 33-40]     |
| `[@Box1973; @Acemoglu2000]` | [@Box1973; @Acemoglu2000] |
| `@Box1973 [33-40]`          | @Box1973 [33-40]          |

## Reference List

A reference list with the full bibliographic details of all cited documents
will be **automatically created** at the end of your document.

Tip: Put a `# References` at the very end of your R Markdown document to
have a section heading before the reference list.

## Citing R and R Packages

Why cite?

- Give **credit** to the software authors (just like when citing literature).

- Enable **reproducible research**: identify which software you used and **which
version**.

## Citing R and R Packages

Base R way: print citation, copy BibTeX  entry into your *.bib* file.

**Cite R**:

```{r}
toBibtex(citation())
```

## Citing R and R Packages

**Cite R Packages**:

```{r}
toBibtex(citation('dplyr'))
```

## Citing R and R Packages: LoadandCite

The dynamic literate programming way: Use `LoadandCite` from the
[repmis](http://christophergandrud.github.io/repmis/) package.

Load all of the packages at the beginning of you R Markdown file in a chunk
with `include=FALSE`.

`LoadandCite` loads the packages and creates a BibTeX file with all of the citations.

```{r, message=FALSE, warning=FALSE}
pkgs <- c('dplyr', 'ggplot2')

repmis::LoadandCite(pkgs, file = 'RpackageCitations.bib')
```

**Note:** Use a file name that is different from your literature BibTeX file!

## Citing R and R Packages: LoadandCite

Include the .bib file in your RMarkdown header.

Each **cite key** follows: `R-PKG_NAME`.

R itself has the key `CiteR`.

So `@R-dplyr` and `@CiteR` create the citations:

- @R-dplyr

- @CiteR

## <i class="fa fa-clock-o"></i> Time Consuming/Intensive Analyses

Knitting your analysis and presentation documents together by placing all of your
R code into code chunks can sometimes be **problematic**:

- When they are **time consuming** (requires a lot of computational time).

- When they **access files over the internet** (bad practice to make many repeated
    calls to the same URL, can crash the site. This is equivalent to a 
    [denial-of-service attack](http://en.wikipedia.org/wiki/Denial-of-service_attack)).

- When they are **many lines long**.

## Solutions

- Long lines: use `source()` to run R code in other files.

- [Caching](http://en.wikipedia.org/wiki/Cache_(computing)) for
time/computationally intensive work: `cache=TRUE` code chunk option: only runs
the chunk when the **chunk code changes**.

## Make files

Make files are the ultimate solution to these problems.

Make is a command line program.

Big Idea: run a *make* file that runs a list of specific files in order.

Files are **only run** if they have been **changed** since the last time the make
file was last run.

See Ch. 6 of RRRR if you might want to do this.

## <i class="fa fa-arrow-circle-o-up"></i> Seminar: Practice

<br>
<br>
<br>

Clone the [HertieDataScience/Examples](https://github.com/HertieDataScience/Examples) repo and play around with **SimplePaperWithAnalysis**.

## <i class="fa fa-arrow-circle-o-up"></i> Seminar: Begin working on your Proposal

Begin working with your partner on your research proposal.

- Identify the research area and key literature.

- Create a new repo and R Markdown document for your proposal.

- Begin building a BibTeX database for your key literature and try including them
in your proposal.

- Begin identifying data sources.

## References
